Xolotl: An Intuitive and Flexible Neuronal Simulator

### 2018-05-25 PaperLatex/paper.tex
Aesthetic fixes and some clarifications

\title[xolotl: neuronal simulator]{Xolotl: An Intuitive and Definitively Sexy Neuronal Simulator}
\author[\firstAuthorLast ]{\Authors}
\address{}
\correspondance{}
\extraAuth{}
\topic{An Intuitive Neuronal Simulator}

@@

	\texttt{xolotl} is an open-source neuronal simulator written in \texttt{C++} with \texttt{MATLAB} wrappers. Complex models and networks can be designed efficiently using an intuitive language tightly coupled to the object-based architecture of the underlying \texttt{C++} code. Models can be specified by adding conductances to compartment objects. The structure is modular, serialized, and searchable, permitting high-level programmatic control over nearly all features of the models. \texttt{C++} templates are provided for developing new conductances, compartments, and integration schemata. It also includes a customizable graphical user interface (GUI, "puppeteer") for rapid prototyping and hand-tuning conductances in real-time. The modular structure and accessibility to all parameters, variables, and dynamics of the model network in \texttt{MATLAB} facilitate rapid construction and assessment of model networks. \texttt{xolotl} is freely available at \url{https://github.com/marderlab/xolotl}. This tool provides straightforward implementation and fast simulation of neuronal models while permitting full control over every aspect of the network and integration.

	\tiny
	\keyFont{ \section{Keywords:} simulator, MATLAB, C++, xolotl, conductance-based, computational, keyword, keyword} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.

@@

\texttt{xolotl} (\url{https://github.com/sg-s/xolotl}) is a fast single-compartment and multi-compartment simulator in \texttt{C++} with \texttt{MATLAB} wrappers (\url{https://www.mathworks.com/products/matlab.html}). Written with an emphasis on flexibility and speed, \texttt{xolotl} can simulate single-compartment conductance-based models, networks of these, and detailed multi-compartment models. \texttt{xolotl} exploits a novel automatic type system, \texttt{cpplab}, which binds \texttt{MATLAB} code to \texttt{C++} header files, creating objects and classes \textit{ad libitum} in \texttt{MATLAB} which reflect the underlying object-oriented code. \texttt{xolotl} implements \texttt{cpplab} to represent the nested structure of conductance-based models, and exploits the computational efficiency of the low-level programming language to quickly integrate models. For this reason, models can be implemented entirely in \texttt{MATLAB} with few lines of code.

Models are specified in \texttt{MATLAB} by a \texttt{xolotl} object which contains compartment objects which themselves contain conductances. Synapses belong to the \texttt{xolotl} object and connect compartments together. The high-level specification supports arbitrarily large network and multi-compartment morphologies.

\texttt{xolotl} provides parameter optimization capabilities through the algorithm-agnostic \texttt{procrustes} toolbox. Any network parameters accessible through the \texttt{xolotl} structure can be optimized using arbitrary algorithms and objective functions on multi-core computers and high-performance computing clusters.

The software has been implemented in \texttt{MATLAB} due to its ease-of-use and popularity among neuroscientists. \texttt{cpplab} provides a powerful backend for specifying and integrating models without relying on the significantly slower and limiting \texttt{MATLAB codegen}. Minimal experience with \texttt{MATLAB} is required to use \texttt{xolotl}, and all equations and integration methods are provided transparently to the end user. No string parsing of equations is required.

\texttt{xolotl} comes packaged with visualization functions, a graphical user interface (GUI) for real-time manipulation of model parameters. Plotting of voltage, intracellular calcium, conductance gating functions, and time constants is provided by built-in \texttt{xolotl} methods. The \texttt{puppeteer} toolbox permits real-time tuning of any network parameters using numerical sliders in a graphical interface which displays the resultant membrane potential and intracellular calcium traces. The ease-of-use of these tools lends them to pedagogical applications and rapid exploration of toy models. This tool aims to simplify the investigation of dynamics of complex neural network models, facilitate collaborative modeling, and complement other tools being developed in the neuroinformatics community.

@@

Using \texttt{xolotl} in \texttt{MATLAB}, users create a \texttt{xolotl} object and populate it with compartments, synapses, and controllers. Each field is a \texttt{cpplab} object constructed by a function call to \texttt{add}. The model is integrated with the \texttt{integrate} function where the membrane potential, intracellular calcium concentration, controller states, intrinsic currents, and synaptic currents can be outputs.

\texttt{xolotl} comes packaged with a library of pre-existing conductance and synapse objects which greatly simplify the task of constructing model neurons. These objects can be referenced by name and added directly to a compartment. Novel conductance dynamics can be easily written by modifying a template header file contained in the \texttt{xolotl} distribution.

@@

The seminal Hodgkin-Huxley model \autocite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952} contains a fast inactivating sodium conductance which promotes spiking, a non-inactivating potassium delayed rectifer, and a passive leak current (\ref{fig:figurehh}A). A compartment, \texttt{HH}, with membrane capacitance $C_m = 10 ~ \mu\mathrm{F/mm^2}$ and surface area $A = 0.1 ~ \mathrm{mm^2}$ can be specified by \ref{fig:figurehh}B. Compartment, conductance, synapse, and controller properties can be specified during the call to the \texttt{add} function, or after construction, using dot-notation in \texttt{MATLAB} (e.g. \texttt{x.HH.Cm}). \ref{fig:figurehh}C shows the \texttt{MATLAB} command prompt after invoking the \texttt{xolotl} object \texttt{x}, displaying the hierarchical structure inherent in conductance-based treatments of neurodynamics.

This model was constructed using conductances from \cite{liuModelNeuronActivitydependent1998} based on electrophysiological recordings from the lobster stomatogastric ganglion \autocite{turrigianoSelectiveRegulationCurrent1995}. In the absence of applied positive current, the model is quiescent and tonically spikes under 0.2 nA of applied current (\ref{fig:figurehh}D). The \texttt{integrate} function takes the applied current as an argument (e.g. "x.integrate(AppliedCurrent)"), so that the \texttt{xolotl} object is agnostic to integration-specific perturbations. The \texttt{plot} function generates voltage and intracellular calcium traces, where the voltage trace is colored by the dominant current. If the membrane potential is increasing, the strongest instantaneous inward current colors the trace. Conversely, if the membrane potential is decreasing, the strongest outward current colors the trace instead. \ref{fig:figurehh}F-I display the results of the \texttt{show} function. Activation and inactivation steady-states and the voltage-dependent time constants of these gating variables describe the conductance dynamics in absence of other channel types.

>>>>

\title[xolotl: neuronal simulator]{Xolotl: An Intuitive and Flexible Neuronal Simulator}
\author[\firstAuthorLast ]{\Authors}
\address{}
\correspondance{}
\extraAuth{}
\topic{An Intuitive Neuronal Simulator}

@@

	\texttt{xolotl} is a free and open-source neuronal simulator written in \texttt{C++} with \texttt{MATLAB} wrappers. Biophysically-detailed models of networks can be designed efficiently using an intuitive language tightly coupled to the object-based architecture of the underlying \texttt{C++} code. Models can be specified by adding conductances to compartment objects. The structure is modular, serialized, and searchable, permitting high-level programmatic control over nearly all features of the models. \texttt{C++} templates are provided for developing new conductances, compartments, and integration schemata. It also includes a customizable graphical user interface (GUI) for rapid prototyping and hand-tuning conductances in real-time. The modular structure and accessibility to all parameters, variables, and dynamics of the model network in \texttt{MATLAB} facilitate rapid construction and assessment of model networks. \texttt{xolotl} is freely available at \url{https://github.com/marderlab/xolotl}. This tool provides straightforward implementation and fast simulation of neuronal models while permitting full control over every aspect of the network and integration.

	\tiny
	\keyFont{ \section{Keywords:} simulator, MATLAB, C++, conductance-based, neuron, network, keyword} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.

@@

\texttt{xolotl} (\url{https://github.com/marderlab/xolotl}) is a fast single-compartment and multi-compartment simulator in \texttt{C++} with \texttt{MATLAB} wrappers. Written with an emphasis on ease-of-use, \texttt{xolotl} can simulate single-compartment conductance-based models, networks of these, and detailed multi-compartment models. \texttt{xolotl} exploits a novel automatic type system, \texttt{cpplab}, which binds \texttt{MATLAB} code to \texttt{C++} header files, creating objects and classes \textit{ad libitum} in \texttt{MATLAB} which reflect the underlying object-oriented code. \texttt{xolotl} implements \texttt{cpplab} to represent the nested structure of conductance-based models, and exploits the computational efficiency of the low-level programming language to quickly integrate models. For this reason, models can be implemented entirely in \texttt{MATLAB} with a few lines of code.

Models are specified in \texttt{MATLAB} by a nested structure. The \texttt{xolotl} object contains compartments which themselves contain conductances. Synapses belong to the \texttt{xolotl} object and connect compartments together. The high-level specification supports arbitrarily large network and multi-compartment morphologies.

%\texttt{xolotl} provides parameter optimization capabilities through the algorithm-agnostic \texttt{procrustes} toolbox. Any network parameters accessible through the \texttt{xolotl} structure can be optimized using arbitrary algorithms and objective functions on multi-core computers and high-performance computing clusters.

The software has been implemented in \texttt{MATLAB} due to its ease-of-use and popularity among neuroscientists. \texttt{cpplab} provides a powerful backend for specifying and integrating models without relying on the significantly slower and limiting \texttt{MATLAB codegen}. While automated \texttt{C++} transpiling from \texttt{MATLAB} using the proprietary \texttt{coder} can drastically improve performance over loops through strong typing and memory pre-allocation, supervenience of \texttt{MATLAB} over \texttt{C++} prevents efficient use of low-level features, such as passing by reference and object-oriented programming.   Minimal experience with \texttt{MATLAB} is required to use \texttt{xolotl}, and all equations and integration methods are provided transparently to the end user. No string parsing of equations is required.

\texttt{xolotl} comes packaged with visualization functions, a graphical user interface (GUI) for real-time manipulation of model parameters. Plotting of voltage, intracellular calcium, conductance gating functions, and time constants is provided by built-in \texttt{xolotl} methods. The GUI permits real-time tuning of any network parameters using numerical sliders in a graphical interface which displays the resultant membrane potential and intracellular calcium traces. The ease-of-use of these tools lends them to pedagogical applications and rapid exploration of toy models. This tool aims to simplify the investigation of dynamics of complex neural network models, facilitate collaborative modeling, and complement other tools being developed in the neuroinformatics community.

@@

Using \texttt{xolotl} in \texttt{MATLAB}, users create a \texttt{xolotl} object and populate it "cpplab" objects which describe compartments, conductances, synapses, and controllers. The model is integrated with the \texttt{integrate} function where the membrane potential, intracellular calcium concentration, controller states, intrinsic currents, and synaptic currents can be outputs.

\texttt{xolotl} comes packaged with a library of pre-existing conductance and synapse objects which greatly simplify the task of constructing model neurons. These objects can be referenced by name and added directly to a compartment. Novel conductance dynamics can be easily written by modifying a template header file contained in the \texttt{xolotl} distribution, or designed entirely from scratch.

@@

The seminal Hodgkin-Huxley model of action potentials in the squid giant axon \autocite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952} contains a fast inactivating sodium conductance ("NaV"), a non-inactivating delayed rectifier ("Kd"), and a passive leak current (\ref{fig:figurehh}A). A compartment, \texttt{HH}, with membrane capacitance ("Cm") and surface area ("A") can be specified by \ref{fig:figurehh}B. Network properties can be set during construction or afterwards using dot-notation in \texttt{MATLAB} (e.g. \texttt{x.HH.Cm}). \ref{fig:figurehh}C shows the \texttt{MATLAB} command prompt after invoking the \texttt{xolotl} object \texttt{x}, displaying the hierarchical structure inherent in conductance-based treatments of neurodynamics.

This model was constructed using conductances from \cite{liuModelNeuronActivitydependent1998} based on electrophysiological recordings from the lobster stomatogastric ganglion \autocite{turrigianoSelectiveRegulationCurrent1995}. In the absence of applied positive current, the model is quiescent. When 0.2 nA is injected, the model tonically spikes (\ref{fig:figurehh}D). The \texttt{integrate} function takes the applied current as an argument (e.g. "x.integrate(AppliedCurrent)"), so that the \texttt{xolotl} object is agnostic to integration-specific perturbations. The \texttt{plot} function generates voltage and intracellular calcium traces, where the voltage trace is colored by the dominant current. If the membrane potential is increasing, the strongest instantaneous inward current colors the trace. Conversely, if the membrane potential is decreasing, the strongest outward current colors the trace instead. \ref{fig:figurehh}F-I display the results of the \texttt{show} function. Activation and inactivation steady-states and the voltage-dependent time constants of these gating variables describe the conductance dynamics in absence of other channel types.
